---
title: "Analytical Skills for Business (WS 2025/26)"
subtitle: "Business Administration (M. A.)"
author: 
    name: "© Benjamin Gross"
    email: benjamin.gross@ext.hs-fresenius.de
    affiliation: 
      - Hochschule Fresenius - University of Applied Science
      - "Email: benjamin.gross@ext.hs-fresenius.de"
      - "Website: https://drbenjamin.github.io"
date: today
date-format: long
abstract: |
    This document holds the course material for the Analytical Skills for Business course in the Master of Business Administration program. It discusses version control systems such as Git and GitHub for efficient team collaboration, offers an overview of no-code and low-code tools for data analytics including Tableau, Power BI, QlikView, makeML, PyCaret, RapidMiner, and KNIME, and introduces key programming languages such as R, Python, and SQL alongside essential programming concepts like syntax, libraries, variables, functions, objects, conditions, and loops. In addition, it covers working with modern development environments, including Unix-like systems, containers, APIs, Jupyter, and RStudio, and sets expectations for project submissions and evaluation.
format:
  html:
    embed-resources: true
    theme: cerulean
    toc: true
    toc-expand: 2
    toc-depth: 2
    number-sections: true
  pdf:
    keep-tex: false
    include-in-header:
      text: |
        \usepackage[noblocks]{authblk}
        \renewcommand*{\Authsep}{, }
        \renewcommand*{\Authand}{, }
        \renewcommand*{\Authands}{, }
        \renewcommand\Affilfont{\small}
    documentclass: scrartcl
    classoption: [onecolumn, oneside, a4paper]
    linkcolor: blue
    urlcolor: blue
    filecolor: magenta
    citecolor: magenta
    colorlinks: true
    margin-left: "1in"
    margin-right: "1in"
    margin-top: "1in"
    margin-bottom: "1in"
    papersize: a4
    fig-cap-location: top
    toc: true
    toc-depth: 5
    number-sections: true
    lof: false
    lot: false
link-citations: true
csl: "./literature/apa.csl"
suppress-bibliography: true
execute:
  enabled: true
  echo: false
---

\clearpage

# Introduction {#sec-intro}

xy

## Implementing version control systems

* git
* GitHub

## Overview on no-code and low-code tools for data analytics

* Tableau
* Power BI
* QlikView
* makeML
* PyCaret
* Rapidminer
* KNIME

## Overview on Programming languages

* R
* Phyton
* SQL

## Elements of programming languages

* Syntax
* libraries
* variables
* functions
* objects
* conditions
* loops

## Development environments

* Unix-like systems
* containers
* APIs 
* Jupyter
* RStudio

# Descriptive statistics

xy

## Measures of centrality, dispersion, and concentration

xy

## Descriptive analytics 

* univariate data
* bivariate data
* multivariate data

## Techniques 

* constructing
* interpreting
* evaluating 
of scores, rankings, metrics, and composite indicators.

## Visualizing and exploration

* categorical
* numerical
* time series data

## Handling messy data

xy

## Association

* measuring the association of variables, including correlation and regression

## Implementing applications 

in the programming language R for practical data analysis.

# Inferential statistics

xy

## Basic concepts of statistical inference

xy

## Quantification of probability through random variables

xy

## Hypothesis testing

xy

## Confidence intervals, p-values, and statistical tests

xy

## Inferential statistics 

in the programming language R, translating theoretical knowledge into practical applications.

# Predictive analytics

xy

## Data mining techniques

xy

## Regression analysis

xy

## Forecasting in predicting future business outcomes

xy

# Literature

All references for this course.

## Essential Readings

```{r essential_bib, results='asis', echo=FALSE, warning=FALSE}
source("literature/refmanager.r")
render_bib("literature/essential_readings.bib")
```

## Further Readings

```{r further_bib, results='asis', echo=FALSE, warning=FALSE}
source("literature/refmanager.r")
render_bib("literature/further_readings.bib")
```
